---
title: "P8131 HW7"
author: "Brian Jo Hsuan Lee"
date: "4/4/2022"
output: pdf_document
---

Load packages 
```{r, message=F}
library(tidyverse)
library(knitr)
library(nlme)
library(lme4)
```

Import data
```{r}
data = read_csv("HW7-politeness_data.csv", col_types = "ffffd")
```

a) 
**EDA**
```{r}
data %>% 
  mutate(
    attitude = factor(attitude, labels = c("Polite", "Informal"))
  ) %>% 
  ggplot(aes(x = gender, y = frequency, fill = gender)) +
  geom_boxplot() +
  facet_grid(cols = vars(attitude)) +
  scale_fill_manual(labels = c("Female", "Male"), values = c("#A13E60", "#3E7BA1")) +
  labs(
    title = "Relationship between Gender/Attitude and Pitch across Scenarios",
    y = "Pitch (frequency)"
  ) +
  theme(
    plot.title = element_text(size = 11, hjust = 0.5),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

\newpage

b) 
**Fit and interpret a random intercept model for the different subjects**

```{r}
# fit a mixed effect model with estimates chosen to optimize the maximum log-likelihood criterion 
lmm1 = lme (frequency ~ gender + attitude, random = ~1 | subject, data = data, method ='ML')
```

The covariance matrix for the pitch `frequency` $i$ of a particular subject is composed of the marginal variances of each population-shared predictor as its diagonal, and the marginal covariances of any two of those predictors in their corresponding entries. The diagonals are all equal, and the non-diagonal entries are all equal for a linear mixed effect model.

In this random intercept model $Y_{ij} = (\beta_o + b_i) + X_{ij}^T\beta + \epsilon_{ij}$, $Y_{ij}$ and $X_{ij}$ are the estimated $i^{th}$ `frequency` \
and its vector of predictors, `genderM` and `attitudeM`, in condition $j$ of a particular subject. \
$b_i \sim N(0, \sigma^2_b)$ is the random `subject`-specific intercept effect for the $i^{th}$ `frequency`, and \
$\epsilon_{ij} \sim N(0, \sigma_b)$ is the within-`subject` error at condition $j$ for the $i^{th}$ `frequency`. \
Note $b_i$ and $\epsilon_{ij}$ are independent, i.e. cov($b_i$, $\epsilon_{ij}$) = 0, cov($\epsilon_{im}$, $\epsilon_{in}$) = 0 \

The covariance matrix for `frequency` $i$ of a subject is derived with equations \
$cov(Y_{im}, Y_{in}) = cov(b_i+\epsilon_{im}, b_i+\epsilon_{in}) = cov(b_i, b_i) + cov(b_i, \epsilon_{in}) + cov(\epsilon_{im}, b_i) + cov(\epsilon_{im}, \epsilon_{in})$ \
$= Var[b_i] + 0 + 0 + 0 = \sigma^2_b$\
for the marginal covariance between `frequency` $i$ pairs under conditions $m$ and $n$ , and \
$Var[Y_{ij}] = Var[b_i+\epsilon_{ij}] = Var[b_i] + Var[\epsilon_{ij}] =  \sigma^2_b + \sigma^2$

```{r}
# obtain the random subject-specific covariance estimate (sigma^2_b)
randeff_cov = as.double(VarCorr(lmm1)[1,1])

# obtain the random population-shared residual variance estimate (sigma^2)
res_var = as.double(VarCorr(lmm1)[2,1])

# build the covariance matrix for a particular subject with the estimates
# where the marginal variance for the subject is the sum of the two values 
pop_pred = c("genderM", "attitudeinf")
cov_y =
  matrix(
    rep(randeff_cov, length(pop_pred)^2),
    nrow = length(pop_pred),
    dimnames = list(pop_pred, pop_pred)
  )
diag(cov_y) = randeff_cov + res_var

kable(cov_y, "simple")
```

The covariance matrix for the fixed effect estimates
```{r}
kable(vcov(lmm1), "simple")
# # or alternatively ...
# lmm1$varFix
```

BLUPs for subject-specific intercepts, which are the random effect coefficients
```{r}
kable(random.effects(lmm1), "simple")
```

Residuals (is there a better way to show the residuals?)
```{r}
data$frequency-fitted(lmm1)
```

\newpage

c)
**Fit a similar random intercept model - but with an interaction term - and compare it with the first model**
```{r}
# fit a mixed effect model, also with estimates chosen to optimize the maximum log-likelihood criterion
lmm2 = lme (frequency ~ gender * attitude, random = ~1 | subject, data = data, method ='ML') 

# compare it with the first model
lmm1_lmm2_pval = anova(lmm2,lmm1)[2, 9]
ifelse(lmm1_lmm2_pval < 0.05, 
       "Reject the null hypothesis and suggest the new model with the interaction term has a better fit on the data",
       "Fail to reject the null hypothesis and suggest the inclusion of the interaction term does not improve the model fit")
```

After comparing the 2 models using the likelihood ratio test, it is concluded that the interaction term for gender and attitude does not create a better fit for modeling pitch, and therefore it is not significantly associated with pitch.

d)
**Fit and interpret a random intercept model for the different subjects and scenarios**
```{r}
# fit a mixed effect model, again, with estimates chosen to optimize the maximum log-likelihood criterion
lmm3 = lmer(frequency ~ gender + attitude + (1 | subject) + (1 | scenario), data = data, REML = F)
```

As before, the covariance matrix for `frequency` $i$ of a particular subject in a scenario is composed of the marginal variances of each population-shared predictor and the marginal covariances of any two of those predictors.\


In this random intercept model $Y_{ij} = (\beta_o + b_{sub, i} + b_{sce, i}) + X_{ij}^T\beta + \epsilon_{ij}$, \
$Y_{ij}$ and $X_{ij}$ are the estimated $i^{th}$ `frequency` and its vector of predictors, `genderM` and `attitudeM`, \
in condition $j$ of a particular subject and scenario, \
$b_{sub, i} \sim N(0, g_{sub})$ is the random `subject`-specific intercept effect for the $i^{th}$ `frequency`, \
$b_{sce, i} \sim N(0, g_{sce})$ is the random `scenario`-specific intercept effect for the $i^{th}$ `frequency`, \
$\epsilon_{ij} \sim N(0, \sigma_b)$ is the within-`subject`-`scenario` error at condition $j$ for the $i^{th}$ `frequency`. \
Note $b_{sub, i}$, $b_{sce, i}$ and $\epsilon_{ij}$ are independent, i.e. $cov(b_{sub, i}, b_{sce, i}) = 0$. \

The covariance matrix for `frequency` $i$ of a particular subject and scenario is derived with equations \
$cov(Y_{im}, Y_{in}) = cov(b_{sub, i}+b_{sce, i}+\epsilon_{im}, b_{sub, i}+b_{sce, i}+\epsilon_{in}) = cov(b_{sub, i}+b_{sce, i}, b_{sub, i}+b_{sce, i})$ \
$= cov(b_{sub, i}, b_{sub, i}) + cov(b_{sce, i}, b_{sub, i}) + cov(b_{sub, i}, b_{sce, i}) + cov(b_{sce, i}, b_{sce, i}) = Var[b_{sub, i}] + 0 + 0 + Var[b_{sce, i}] = g_{sub}  + g_{sce}$ \
for the marginal covariance between `frequency` $i$ pairs under conditions $m$ and $n$, and \
$Var[Y_{ij}] = Var[b_{sub, i}+b_{sce, i}+\epsilon_{ij}] = Var[b_{sub, i}] + Var[b_{sce, i}] + Var[\epsilon_{ij}] = g_{sub} + g_{sce} + \sigma^2$

```{r}
cov_obj = as.data.frame(VarCorr(lmm3))

# obtain the residual variance estimate (sigma^2)
res_var2 = cov_obj[3,4]

# obtain the subject covariance estimate (sigma^2_b1)
sub_cov = cov_obj[2,4]

# obtain the scenario covariance estimate (sigma^2_b1)
sce_cov = cov_obj[1,4]

# build a covariance matrix with the covariance and variance estimates
# where the variance for Y is the sum of the two values
cov_y2 =
  matrix(
    rep(sub_cov + sce_cov, length(pop_pred)^2),
    nrow = length(pop_pred),
    dimnames = list(pop_pred, pop_pred)
  )
diag(cov_y2) = sub_cov + sce_cov + res_var2

kable(cov_y2, "simple")
```

